---
title: "Exploratory Data Analysis Project: Effect of Tobacco Exposure on Child"
author: "Angel Zheng"
date: "Oct 2023 for PHP2550"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE)

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse) # for data manimulation
library(ggmice) # for missing pattern
library(naniar) # for missing summary
library(kableExtra) # for neat output
library(gtsummary) # for neat summary table 
library(ggpubr) # for combining ggplots

```

# Data Background
```{r, echo=FALSE}
tobacco_exposure <-  read.csv("C:/ANGEL/Brown 22-24/23Fall/PHP2550_pda/datasets/project1.csv")

```

Tobacco exposure plays a significant role in behavioral and child developmental research, particularly concerning smoking during pregnancy (SDP) and environmental tobacco smoke (ETS) post-delivery, both potentially linked to children's health and behavior throughout their lives. This report aims to explore the correlation between SDP and ETS with externalizing behaviors, self-regulation, and substance use.

The dataset utilized for this report stems from a follow-up study conducted by Dr. Micalizzi. This study focuses on a subset of adolescents (N=100) and their mothers, randomly selected for participation from a previous investigation centered on smoke avoidance intervention among low-income women to reduce SDP and ETS.

[1]The original study recruited approximately 800 pregnant individuals, all exposed to smoke (including current smokers, those who quit smoking on their own, or those exposed to secondhand smoke), expecting a single baby, and having access to a telephone and video player, were randomly assigned to either an experimental or control group. Those in the experimental group received personalized newsletters focused on quitting smoking (5 during pregnancy, 3 post-pregnancy), and custom-tailored videos based on behavioral surveys (3 during pregnancy, 2 post-pregnancy). Meanwhile, the control group received general newsletters and videos covering various aspects of healthy pregnancy.

Key areas of focus within this report include demographic information of both parents and children, variables examining their relationships, factors related to externalizing behaviors (EXT), aspects of self-regulation (SR), and variables pertaining to substance use (SU).

Externalizing behaviors encompass three main aspects: assessment via the Brief Problem Monitor (BPM) questionnaire, the SWAN rating, and evaluation for Autism diagnosis.

Regarding self-regulation, the study emphasizes emotion regulation, specifically assessed using the Emotion Regulation Questionnaire, focusing on Cognitive Reappraisal and Expressive Suppression.

Lastly, the dataset records information on children's usage of cigarettes, e-cigarettes, marijuana, and alcohol, representing primary and prevalent substance-use issues among adolescents.

```{r, echo=FALSE}
#colnames(tobacco_exposure)

```

# Data Preprocessing
The processing began by filtering child data to select specific columns related to the child baseline arm. Demographic variables were chosen, excluding unnecessary columns and renaming certain variables for clarity. Subsequently, sections of the dataset pertaining to cigarette, e-cigarette, marijuana, and alcohol usage were modified to create summarized variables for further analysis.

The data manipulation continued by addressing the Brief Problem Monitor scoring and emotional regulation aspects, aggregating and selecting specific columns accordingly. Some sections of the dataset, such as physical development scale, life stress assessments, and variables related to dysregulation, were dropped to simplify the analysis.

Similarly, alterations were made to the parent data, focusing on demographic variables and scoring related to parental monitoring, while excluding redundant or unrelated columns. Scoring for parental knowledge, child disclosure, parental solicitation, and control were derived based on specific columns. Sections concerning chaos, adult temperament, and stress were removed for the purpose of the research.

Upon thorough examination of the dataset, various anomalies were detected, ranging from formatting inconsistencies to substantial outliers that indicated potential errors. To address these discrepancies, several corrective measures were undertaken during the data preprocessing phase. Instances of numeric entries, for instance '250,000,' were standardized to '250000' for consistency in representation. Outliers, such as responses exceeding the maximum allowable value, for example '40' where the maximum value should be '30,' were identified and marked as 'NA' to flag potential inaccuracies. In cases where entries represented ranges, such as '20-25,' the mean was computed and rounded to the nearest integer, ensuring coherence in the dataset. Additionally, meticulous checks were conducted on binary variables, unifying their format to ensure consistency across all records. These meticulous steps were pivotal in rectifying anomalies and establishing uniformity in the dataset for subsequent analyses.

Details of how those entries are modified are in code appendix. 
```{r, echo=FALSE}
# make "250,000" as "250000" to uniform numeric format
tobacco_exposure$income[which(tobacco_exposure$income == "250, 000")] <- 250000 
tobacco_exposure$income <- as.integer(tobacco_exposure$income)

# Null entry with 40 days of smoking in 30 days
tobacco_exposure$momcig[which(tobacco_exposure$momcig == 40)] <- NA

# Null entry with unreasonable response
tobacco_exposure$mom_numcig[which(tobacco_exposure$mom_numcig == 44989)] <- NA

# convert text response to numeric
tobacco_exposure$mom_numcig[which(tobacco_exposure$mom_numcig == "2 black and miles a day")] <- 2
tobacco_exposure$mom_numcig[which(tobacco_exposure$mom_numcig == "20-25")] <- 23
tobacco_exposure$mom_numcig[which(tobacco_exposure$mom_numcig == "None")] <- 0
tobacco_exposure$mom_numcig <- as.integer(tobacco_exposure$mom_numcig)

```

```{r, echo=FALSE}
# extract numeric part from string responses
string_to_binary <- function(string_var) {
  string_num <- ifelse(string_var == "", NA, # NA if blank
                       as.integer(gsub("([0-9]+).*$", "\\1", string_var)))
  string_binary <- replace(string_num, string_num == 2, 0)
  return(string_binary)
}

tobacco_exposure[, c("mom_smoke_16wk", "mom_smoke_22wk", "mom_smoke_32wk",
                   "mom_smoke_pp1", "mom_smoke_pp2",
                   "mom_smoke_pp12wk", "mom_smoke_pp6mo")] <- lapply(
                     tobacco_exposure[, c("mom_smoke_16wk", "mom_smoke_22wk", "mom_smoke_32wk",
                                        "mom_smoke_pp1", "mom_smoke_pp2",
                                        "mom_smoke_pp12wk", "mom_smoke_pp6mo")], 
                     string_to_binary)

```

```{r, echo=FALSE}
# correct 0 in SWAN response to NA
tobacco_exposure$swan_inattentive[which(tobacco_exposure$swan_inattentive == 0)] <- NA
tobacco_exposure$swan_hyperactive[which(tobacco_exposure$swan_hyperactive == 0)] <- NA

```

# Demographic Information, univariate
The demographic information is displayed below. The participants contain parents with race of white the most (61%), with no Asians in both parents and children. There is a relative balance in Hispanic/Latino ratios. However, with the imbalance in race, it is worth noting that the study might have limited generalizability. 
```{r, echo=FALSE}
# create a datafrmae for reporting race and ethnicity 
race_ethnicity <- 
  tobacco_exposure %>% 
  mutate(
    parent_ethnicity = case_when(
      pethnic == 1 ~ "Hispanic or Latino",
      pethnic == 0 ~ "None Hispanic nor Latino", 
      pethnic == 2 ~ "Prefer not to answer"),
    parent_race = case_when(
      paian == 1 ~ "American Indian/Alaska Native",
      pasian == 1 ~ "Asian",
      pnhpi == 1 ~ "Native Hawaiian/Pacific Islander",
      pblack == 1 ~ "Black",
      pwhite == 1 ~ "White", 
      prace_other == 1 ~ "Others"),
    child_ethnicity = case_when(
      tethnic == 1 ~ "Hispanic or Latino",
      tethnic == 0 ~ "None Hispanic nor Latino", 
      tethnic == 2 ~ "Prefer not to answer"),
    child_race = case_when(
      taian == 1 ~ "American Indian/Alaska Native",
      tasian == 1 ~ "Asian",
      tnhpi == 1 ~ "Native Hawaiian/Pacific Islander",
      tblack == 1 ~ "Black",
      twhite == 1 ~ "White", 
      trace_other == 1 ~ "Others")
  ) %>%
  select(parent_ethnicity, parent_race, child_ethnicity, child_race) 

# Table 1: summary of race and ethnicity 
data.frame(parent_child = rep(c("Parent", "Child"), each = 49),
           ethnicity = c(race_ethnicity$parent_ethnicity, race_ethnicity$child_ethnicity),
           race = c(race_ethnicity$parent_race, race_ethnicity$child_race)) %>%
  tbl_summary(by = parent_child, missing = "no") %>% 
  modify_header(label ~ "**Variable**") %>%
  as_kable_extra(booktabs = TRUE,
                 caption = "Ethnicity and Race Demographic",
                 longtable = TRUE) %>%
  kableExtra::kable_styling(font_size = 8,
                            latex_options = c("repeat_header", "HOLD_position"))

```

Table 2 shows the employment status, highest education, and child sex. Most parents have full-time job and above-highschool degree (`pedu`> 1). The mean and median of parent income are observably different, indicating that the income is skewed towards lower side with more parents with lower income. The histogram of parent and child age is also being looked at in Figure 1, but it does not seem problematic. 
```{r, echo=FALSE}    
# Table 2: Summary of socioeconomic and child sex
tobacco_exposure %>%
  select(c(12:14,2,52:53)) %>%
  tbl_summary(missing="no",
              type = all_continuous() ~ "continuous2",
              statistic = list(all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})")))  %>% 
  modify_header(label ~ "**Variable**") %>%
  as_kable_extra(booktabs = TRUE,
                 caption = "Socioeconomic and Child Sex Demographic",
                 longtable = TRUE) %>%
  kableExtra::kable_styling(font_size = 8,
                            latex_options = c("repeat_header", "HOLD_position"))

```


# Missing Pattern
This dataset is problematic in extent of missingness. Below is the table showing percentage missingness for the variables, and it can be seen that `num_30` of all substances is majority missing. This is due to that this variable is only collected if the child reported "yes" to the previous variable. Therefore, the `num_30` variables are excluded from this analysis; instead, we use the previous binary responses to substances use. 

`mom_smoke_pp1` and `mom_smoke_pp2` also have a large proportion missing. Looking at the data it could be due to loss of followup after delivery of baby. Therefore, for uniformity, we focus on the latter smoke exposure report instead of the followups of self-report mom smoke for postnatal tobacco exposure.  

```{r, echo=FALSE}
# Missingness summary for all variables
missingness_te <- 
  tobacco_exposure %>% 
  miss_var_summary() %>%
  filter(n_miss > 0) 

colnames(missingness_te) <- c("Variable Name", "Variable", "Number of Missing", "Percentage % of Missing")
  
missingness_te %>%
  kable(booktabs = TRUE, caption = "Missingness Summary", longtable = TRUE) %>%
  kableExtra::kable_styling(font_size = 8, latex_options = c("repeat_header", "HOLD_position"))

```


```{r, echo=FALSE}
# convert cotinine level to categorical
cotinine_pre <- 
  ifelse(tobacco_exposure$cotimean_34wk < 10, 0,
       ifelse(tobacco_exposure$cotimean_34wk < 30, 1, 2))
       
cotinine_post <- 
  ifelse(tobacco_exposure$cotimean_pp6mo < 10, 0,
       ifelse(tobacco_exposure$cotimean_pp6mo < 30, 1, 2))

# a dataframe for prenatal responses to self-report smoking behavior
prenatal_severity <- 
  tobacco_exposure[,c("parent_id", "mom_smoke_16wk", "mom_smoke_22wk", "mom_smoke_32wk")] %>%
  mutate(cotinine = cotinine_pre,
         prenatal_score = as.factor(cotinine + mom_smoke_16wk + mom_smoke_22wk + mom_smoke_32wk))

# a dataframe for postnatal responses to tobacco exposure
postnatal_severity <- 
tobacco_exposure[,c("parent_id", "smoke_exposure_6mo", "smoke_exposure_12mo", 
                    "smoke_exposure_2yr", "smoke_exposure_3yr", "smoke_exposure_4yr", "smoke_exposure_5yr")] %>%
  mutate(cotinine = cotinine_post,
         postnatal_score = as.factor(cotinine + smoke_exposure_6mo + smoke_exposure_12mo + 
                    smoke_exposure_2yr + smoke_exposure_3yr + smoke_exposure_4yr + smoke_exposure_5yr))
```


# Prenatal and Postnatal Smoking
In the below two tables, it can be seen that at each timepoint for both prenatal and postnatal period, the mothers who reported smoking or smoking exposure of children are the relative minority (around 30%). 
```{r, echo=FALSE}
# Table summary of prenatal smoking 
tbl_summary(prenatal_severity[,-1], missing = "no") %>% 
  modify_header(label ~ "**Variable**") %>%
  as_kable_extra(booktabs = TRUE,
                 caption = "Prenatal Tobacco Exposure (Smoking During Pregnancy)",
                 longtable = TRUE) %>%
  kableExtra::kable_styling(font_size = 8,
                            latex_options = c("repeat_header", "HOLD_position"))

# Table 5: summary of postnatal smoking
tbl_summary(postnatal_severity[,-1], missing = "no") %>% 
  modify_header(label ~ "**Variable**") %>%
  as_kable_extra(booktabs = TRUE,
                 caption = "Postnatal Tobacco Exposure (Environmental Tobacco Smoke)",
                 longtable = TRUE) %>%
  kableExtra::kable_styling(font_size = 8,
                            latex_options = c("repeat_header", "HOLD_position"))

```

```{r, echo=FALSE}
SDP_ETS <- na.omit(merge(prenatal_severity, postnatal_severity, by = "parent_id"))

```

This figure shows an relationship between SDP summary scores and ETS summmary scores, which are computed by adding the smoking behavior and cotinine levels. [2]Cotinine are categorized into 3 levels: nonsmoker (0) with cotinine < 10, passive-smoker (1) with cotinine < 30, and active smoker (2) with cotinine > 30 ng/mL. In the bar chart, it is seen that at lower ETS scores, there are more subjects with low SDP scores; and at higher ETS scores, there are mostly high SDP scores subjects. This demonstrates that SDP and ETS are related to each other positively, and is also confirmed by their correlation value $r = 0.683$ which is nearly highly correlated with each other. 

```{r, fig.height=3, fig.cap="The Relationship Between SDP and ETS Summary Scores", echo=FALSE}
ggplot(SDP_ETS, aes(postnatal_score)) + 
  geom_bar(aes(fill=prenatal_score)) +
  xlab("ETS Score") +
  guides(fill=guide_legend(title="SDP Score"))

```

```{r, echo=FALSE}
#cor(as.numeric(SDP_ETS$prenatal_score), as.numeric(SDP_ETS$postnatal_score))

```


# Externalizing Behaviors vs. SDP/ETS
Recall that autism diagnosis is a part of externalizing behaviors, but it won't be included because as only 1 child has been diagnosed and 1 child is suspected, while all others no autism or missing response. In this case, including autism would not be helpful to the analysis. 

```{r, echo=FALSE}
# a subset of dataframe selecting variables of externalizing
externalizing <- 
  tobacco_exposure %>%
  select("parent_id", "bpm_att", "bpm_att_p", "bpm_ext", "bpm_ext_p", "swan_hyperactive", "swan_inattentive", "childasd") %>%
  mutate(bpm_att_mean = (bpm_att + bpm_att_p)/2) %>% # child attention score from self and parent
  mutate(bpm_ext_mean = (bpm_ext + bpm_ext_p)/2) %>% # child externalizing score from self and parent
  merge(prenatal_severity, "parent_id") %>% # combine with prenatal exposure 
  merge(postnatal_severity, "parent_id")  # combine with postnatal exposure 

```

```{r, echo=FALSE}
# autism data distribution
#table(externalizing$childasd, useNA = "always") # not helpful to include autism data

```

```{r, echo= FALSE}
bpm_colors <- c("BPM Attention" = "#F8766D", "BPM Externalizing" = "#7CAE00")

BPM_SDP <- 
  ggplot(subset(externalizing, !is.na(prenatal_score)), aes(prenatal_score)) + 
  geom_jitter(aes(y=bpm_att_mean, color = "BPM Attention"), width = 0.2) +
  geom_jitter(aes(y=bpm_ext_mean, color = "BPM Externalizing"), width = 0.2) +
  labs(x = "SDP", y = "BPM Scores", color = "BPM") +
  scale_color_manual(values = bpm_colors)

BPM_ETS <-
  ggplot(subset(externalizing, !is.na(postnatal_score)), aes(postnatal_score)) + 
  geom_jitter(aes(y=bpm_att_mean, color = "BPM Attention"), width = 0.2) +
  geom_jitter(aes(y=bpm_ext_mean, color = "BPM Externalizing"), width = 0.2) +
  labs(x = "ETS", y = "BPM Scores", color = "BPM") +
  scale_color_manual(values = bpm_colors)

```

```{r, fig.height=3, fig.cap="Effect of SDP/ETS on BPM Scores", echo=FALSE}
ggarrange(BPM_SDP, BPM_ETS, common.legend = TRUE)

```

BPM and SWAN questionnaires are plotted with different colors.

```{r, echo=FALSE}
swan_colors <- c("SWAN Hyperactive" = "#00BFC4", "SWAN Inattentive" = "#C77CFF")

SWAN_SDP <-
  ggplot(subset(externalizing, !is.na(prenatal_score)), aes(prenatal_score)) + 
  geom_jitter(aes(y=swan_hyperactive, color = "SWAN Hyperactive"), width = 0.2) +
  geom_jitter(aes(y=swan_inattentive, color = "SWAN Inattentive"), width = 0.2) +
  labs(x = "SDP", y = "SWAN Scores", color = "SWAN") +
  scale_color_manual(values = swan_colors)

SWAN_ETS <-
  ggplot(subset(externalizing, !is.na(postnatal_score)), aes(postnatal_score)) + 
  geom_jitter(aes(y=swan_hyperactive, color = "SWAN Hyperactive"), width = 0.2) +
  geom_jitter(aes(y=swan_inattentive, color = "SWAN Inattentive"), width = 0.2) +
  labs(x = "ETS", y = "SWAN Scores", color = "SWAN") +
  scale_color_manual(values = swan_colors)

```


```{r, fig.height=3, fig.cap="Effect of SDP/ETS on SWAN Scores", echo=FALSE}
ggarrange(SWAN_SDP, SWAN_ETS, common.legend = TRUE)

```

The figures display the relationship between SDP/ETS and the questionnaire response on externalizing behaviors, specifically the BPM and SWAN scores. Regardless of the limited data points, it can still be seen that as the severity of SDP and ETS increases, the questionnaire responses tend to cluster at a higher score. This suggests a sign that SDP and ETS is associated with increased problem in externalizing behaviors. 

In this section, some correlations are also noted: 

`cor(Attention BPM Score of Child, Attention BPM Score of Parent)` $r=0.571$

`cor(Severity of Postnatal Smoke Exposure, Child-Rated Parental Knowledge)` $r=-0.189$

`cor(Severity of Postnatal Smoke Exposure, Child-Rated Child Disclosure)` $r=-0.290$

BPM score of child and parent are moderately correlated ($r=0.571$), which adds a confounding to the associations between attention BPM scores and SDP/ETS. 

Also, severtiy of postnatal ETS is negatively correlated to both child-rated parental knowledge and child disclosure, which draws a relationship between ETS and irresponsibility of parent (although the correlation is not strong). This provides a counter-argument to the validity of SWAN rating, which was by the parent to the child. In the families with more severe postnatal ETS, the parents might not be responsible enough to know if their children are showing symptoms for ADHD. 

```{r, eval=FALSE, echo=FALSE}
# some correlations being tested 
cor(externalizing$bpm_att_mean, tobacco_exposure$bpm_att_a, use = "pairwise.complete.obs")
cor(externalizing$bpm_ext_mean, tobacco_exposure$bpm_ext_a, use = "pairwise.complete.obs")
cor(postnatal_severity$postnatal_score, tobacco_exposure$pmq_parental_knowledge, use = "pairwise.complete.obs")
cor(postnatal_severity$postnatal_score, tobacco_exposure$pmq_child_disclosure, use = "pairwise.complete.obs")

```


# Self Emotion Regulation vs. SDP/ETS
```{r, echo=FALSE}
# prepare format for plotting
erq.pre <- prenatal_severity %>%
  merge(tobacco_exposure[,c("parent_id", "erq_cog", "erq_exp")], "parent_id") %>%
  select(-prenatal_score) %>%
  pivot_longer(cols = starts_with("mom_smoke_"), names_to = "timepoint", values_to = "smoke") %>%
  subset(!is.na(smoke)) 

erq.pre$timepoint <- sub("^mom_smoke_(.*)$", "\\1", erq.pre$timepoint)

```


In this section, because the outcome is still responses to questionnaires, we take similar approach as in the last section. Boxplots and scatter points here show ERQ scores on cognitive reappraisal and expressive suppression do not show a clear relationship with SDP or ETS. The median is varying, and the range is also drastically different. Boxplots are included to demonstrate that at some levels of SDP and ETS, there were extremely limited entries which could be the reason that this relationship is difficult to examine. 

```{r, echo=FALSE}
erq.pre <-
  prenatal_severity %>%
  merge(tobacco_exposure[,c("parent_id", "erq_cog", "erq_exp")], "parent_id") %>%
  select(6:8) %>%
  pivot_longer(cols=!prenatal_score, names_to = "ERQ", values_to = "Score")

erq.post <-
  postnatal_severity %>%
  merge(tobacco_exposure[,c("parent_id", "erq_cog", "erq_exp")], "parent_id") %>%
  select(9:11) %>%
  pivot_longer(cols=!postnatal_score, names_to = "ERQ", values_to = "Score")

```

```{r, echo=FALSE}
ERQ_SDP <-
  ggplot(aes(x=prenatal_score, y=Score), data=subset(erq.pre, !is.na(prenatal_score))) + 
  geom_boxplot(aes(fill=ERQ), alpha=0.5) +
  geom_jitter(aes(color=ERQ), width=0.2) +
  scale_fill_brewer(palette = "Pastel1") +  
  scale_color_brewer(palette = "Pastel1") +
  labs(x="SDP")

ERQ_ETS <- 
  ggplot(aes(x=postnatal_score, y=Score), data=subset(erq.post, !is.na(postnatal_score))) + 
  geom_boxplot(aes(fill=ERQ), alpha=0.5) +
  geom_jitter(aes(color=ERQ), width=0.2) +
  scale_fill_brewer(palette = "Pastel1") +  
  scale_color_brewer(palette = "Pastel1") +
  labs(x="ETS")

```

```{r, fig.height=3, fig.cap= "Effect of SDP/ETS on Self Emotion Regulation", echo=FALSE}
ggarrange(ERQ_SDP, ERQ_ETS, common.legend = TRUE)

```

In this section, some correlations are also noted: 

`cor(ERQ Score of Child Cognitive Reappraisal, ERQ Score of Child Expressive Suppression)` $r=0.119$

`cor(ERQ Score of Child Cognitive Reappraisal, ERQ Score of Parent Cognitive Reappraisal)` $r=0.090$

`cor(ERQ Score of Child Expressive Suppression, ERQ Score of Parent Expressive Suppression)` $r=0.258$

The inter-correlation between the two aspects of emotion control is weak. The correlations between parent's emotion control and child's are also weak. 

```{r, eval=FALSE, echo=FALSE}
cor(tobacco_exposure$erq_cog, tobacco_exposure$erq_exp, use = "pairwise.complete.obs")
cor(tobacco_exposure$erq_cog, tobacco_exposure$erq_cog_a, use = "pairwise.complete.obs")
cor(tobacco_exposure$erq_exp, tobacco_exposure$erq_exp_a, use = "pairwise.complete.obs")

```

# Substance-use vs. Prenatal and Postnatal Period Tobacco Exposure
```{r, echo=FALSE}
# prepare format for plotting
substance.pre <- 
  tobacco_exposure %>%
  select("parent_id", "cig_ever", "e_cig_ever", "mj_ever", "alc_ever") %>%
  merge(prenatal_severity, "parent_id") %>%
  pivot_longer(cols = starts_with("mom_smoke_"), names_to = "timepoint", values_to = "smoke") %>%
  subset(!is.na(prenatal_score)) %>%
  group_by(timepoint, smoke) %>%
  mutate(cig_prop = sum(cig_ever, na.rm = TRUE)/n(),
         e_cig_prop = sum(e_cig_ever, na.rm = TRUE)/n(),
         mj_prop = sum(mj_ever, na.rm = TRUE)/n(),
         alc_prop = sum(alc_ever, na.rm = TRUE)/n())

# modify timepoint format
substance.pre$timepoint <- sub("^mom_smoke_(.*)$", "\\1", substance.pre$timepoint)

# labels for facets
substance.labs <- c("Alcohol", "Cigarettes", "E-Cigarettes", "Marijuana")
names(substance.labs) <- c("alc_prop", "cig_prop", "e_cig_prop", "mj_prop")

```

```{r, echo=FALSE}
# prepare format for plotting
substance.post <- 
  tobacco_exposure %>%
  select("parent_id", "cig_ever", "e_cig_ever", "mj_ever", "alc_ever") %>%
  merge(postnatal_severity, "parent_id") %>%
  pivot_longer(cols = starts_with("smoke_exposure"), names_to = "timepoint", values_to = "smoke") %>%
  subset(!is.na(postnatal_score)) %>%
  group_by(timepoint, smoke) %>%
  mutate(cig_prop = sum(cig_ever, na.rm = TRUE)/n(),
         e_cig_prop = sum(e_cig_ever, na.rm = TRUE)/n(),
         mj_prop = sum(mj_ever, na.rm = TRUE)/n(),
         alc_prop = sum(alc_ever, na.rm = TRUE)/n())

# modify timepoint format  
postnatal_order <- c("6mo", "12mo", "2yr", "3yr", "4yr", "5yr")
substance.post$timepoint <- factor(sub("^smoke_exposure_(.*)$", "\\1", substance.post$timepoint), levels=postnatal_order)

```

```{r, echo=FALSE}
# bar plots for substance use vs. prenatal smoke status
substance_bar.pre <- 
  substance.pre %>%
  select(timepoint, smoke, cig_prop:alc_prop) %>%
  unique() %>%
  gather(substance, prop, cig_prop:alc_prop) %>%
  ggplot(aes(x=timepoint, y=prop, fill=as.factor(smoke))) + 
    geom_col(position="dodge") + 
    facet_grid(. ~ substance, labeller = labeller(substance=substance.labs)) +
    xlab("prenatal timepoints") + ylab("proportion of children with substance-use") +
    scale_fill_discrete(name = "Smoke Exposure Status", labels = c("Non Smoker/No Exposure", "Smoker/Exposure"))

# bar plots for substance use vs. postnatal smoke status
substance_bar.post <- 
substance.post %>%
  select(timepoint, smoke, cig_prop:alc_prop) %>%
  unique() %>%
  gather(substance, prop, cig_prop:alc_prop) %>%
  ggplot(aes(x=timepoint, y=prop, fill=as.factor(smoke))) + 
    geom_col(position="dodge") +
    facet_grid(. ~ substance, labeller = labeller(substance=substance.labs)) +
    xlab("postnatal timepoints") + ylab("proportion of children with substance-use")

```

There is relatively less available data to analyze in this substance-use section. Relating to the age of children being included, it is reasonable that some of them never tried any of the substance. 

This section uses bar charts with proportion of children using certain substance. Color of the bar indicates the smoking or exposure status. In the prenatal barplots, for all timepoints, there is a higher proportion of substance-use child for all types of substance if the mother self-reported as smoker (Missing bar due to proportion = 0, no child with subtance use in that group). However, in the postnatal barplots, there is higher proportion of cigarettes and marijuanause for children with non-smoker mother.

```{r, echo=FALSE, fig.cap="Substance Use vs. Prenatal Timepoints by Substance Type", fig.height=3}
# FSubstance Use vs. Prenatal Timepoints by Substance Type
ggarrange(substance_bar.pre, substance_bar.post, common.legend = TRUE, legend="bottom")

```

```{r, eval=FALSE, echo=FALSE}
# substance use proportion vs. prenatal cotinine levels of parent
# not included in report, discarded 
substance.pre %>%
  merge(tobacco_exposure[,c("parent_id", "cotimean_34wk")], "parent_id") %>%
  gather(substance, prop, cig_prop:alc_prop) %>%
  select(9:11) %>%
  unique() %>%
  ggplot(aes(x=cotimean_34wk, y=prop, color=substance)) + 
    geom_point() + geom_smooth(method = lm, se = FALSE)

# substance use proportion vs. postnatal cotinine levels of baby
# not included in report, discarded 
substance.post %>%
  merge(tobacco_exposure[,c("parent_id", "cotimean_pp6mo_baby")], "parent_id") %>%
  gather(substance, prop, cig_prop:alc_prop) %>%
  select(9:11) %>%
  unique() %>%
  ggplot(aes(x=cotimean_pp6mo_baby, y=prop, color=substance)) + 
    geom_point() + geom_smooth(method = lm, se = FALSE)

```

# Initial Regression
In this section, regressions for the outcome variables in each aspect (EXT, SR, SU) -> `pre/post timepoints`*`smoke status` are generated. 

In the externalizing part, the smoke status in prenatal period is statistically significant $p=0.040$ to the BPM score of attention problems Also, the smoke status in postnatal period is statistically significant $p=0.011$ to the BPM score of externalizing problems. There is no shown significance for the emotion control outcomes. And possibly due to the limited availability of substance use outcomes, the smoke status and timepoints all have small p-values in pre&postnatal periods. 

```{r, eval=FALSE}
# example chunck for initial regression
summary(lm(bpm_att_mean ~ timepoint*smoke, data=ext.pre)) # smoke status significant
summary(lm(bpm_ext_mean ~ timepoint*smoke, data=ext.post)) # smoke status significant

```

```{r, eval=FALSE, echo=FALSE}
# Regressions to check p-values
summary(lm(bpm_att_mean ~ timepoint*smoke, data=ext.pre)) # smoke status significant
summary(lm(bpm_ext_mean ~ timepoint*smoke, data=ext.pre))
summary(lm(swan_hyperactive ~ timepoint*smoke, data=ext.pre))
summary(lm(swan_inattentive ~ timepoint*smoke, data=ext.pre))

summary(lm(bpm_att_mean ~ timepoint*smoke, data=ext.post)) 
summary(lm(bpm_ext_mean ~ timepoint*smoke, data=ext.post)) # smoke status significant
summary(lm(swan_hyperactive ~ timepoint*smoke, data=ext.post))
summary(lm(swan_inattentive ~ timepoint*smoke, data=ext.post))

summary(lm(erq_cog ~ timepoint*smoke, data=erq.pre))
summary(lm(erq_exp~ timepoint*smoke, data=erq.pre))

summary(lm(erq_cog ~ timepoint*smoke, data=erq.post))
summary(lm(erq_exp~ timepoint*smoke, data=erq.post))

summary(lm(cig_prop ~ timepoint*smoke, data=substance.pre)) # smoke and time trend significant
summary(lm(e_cig_prop ~ timepoint*smoke, data=substance.pre))
summary(lm(mj_prop ~ timepoint*smoke, data=substance.pre)) 
summary(lm(alc_prop ~ timepoint*smoke, data=substance.pre)) 

summary(lm(cig_prop ~ timepoint*smoke, data=substance.post)) # smoke and time trend significant
summary(lm(e_cig_prop ~ timepoint*smoke, data=substance.post))
summary(lm(mj_prop ~ timepoint*smoke, data=substance.post)) 
summary(lm(alc_prop ~ timepoint*smoke, data=substance.post))

```

# Conclusion 
After conducting an in-depth Exploratory Data Analysis, the impact of SDP/ETS on adolescent behavior, externalizing tendencies, self-regulation, and substance use was thoroughly investigated. The evidence suggests a noticeable correlation between the presence of SDP/ETS during prenatal or postnatal periods and the onset of externalizing problems in children. These problems encompass attention issues and symptoms resembling ADHD. Moreover, a higher incidence of substance use problems was observed among children with mothers who reported SDP. However, clear evidence regarding the effects of SDP/ETS on emotion regulation remains elusive.

This dataset boasts several strengths and limitations. It encompasses a wide array of variables, including demographic information and multiple measurements of behavioral aspects. This comprehensive range of variables opens up diverse possibilities for data analysis and offers transparency in subject traits for future researchers seeking replication.

Nevertheless, a drawback of this extensive variable range is the presence of missing data. Loss of follow-up and missing baseline variables contribute to this missingness, reducing the dataset's statistical power. Imputation methods aren't suitable due to the small sample size. Furthermore, the limited number of instances with reported SDP/ETS within this small sample size restricts in-depth analysis, potentially leading to chance effects.

Certain variables, notably maternal responses regarding smoking during pregnancy and child-reported substance use, rely on self-reporting and may be influenced by social desirability bias. Subjects might choose not to disclose truthfully, aligning their responses with societal norms.

In conclusion, while the plots in this report serve as a preliminary exploration, a more detailed analysis is advisable before arriving at a conclusive scientific inference. The findings presented here should be viewed as indicative and require further scrutiny.

\newpage
# 6 Reference

[1] Risica, P. M., Gavarkovs, A., Parker, D. R., Jennings, E., & Phipps, M. (2017). A tailored video intervention to reduce smoking and environmental tobacco exposure during and after pregnancy: Rationale, design and methods of Baby's Breath. Contemporary clinical trials, 52, 1–9. https://doi.org/10.1016/j.cct.2016.10.010 

[2] University of Rochester Medical Center. (n.d.). Nicotine & Cotinine. Retrieved from https://www.urmc.rochester.edu/encyclopedia/content.aspx?contentid=nicotine_cotinine&contenttypeid=167#:~:text=Cotinine%20levels%20in%20a%20nonsmoker,more%20than%20500%20ng%2FmL. 

\newpage
# 7 Code Appendix: All code for this report
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
